---
title: "applies"
author: "Timothy Chen Allen"
date: "Monday, May 11, 2015"
output: html_document
---

Use "apply" to apply a function to every row of a matrix
```{r}
# Apply works on each row of a matrix.  Create the matrix first.
# 1) Create a vector (which is "dimensionless"; it's a single
# column of data with many rows.
ROLLS <- 1000000
v <- sample(x=1:6, size = ROLLS, replace = TRUE)
head(v, 20)
# 2) The "matrix" command puts dimensions on this vector.
mx <- matrix(data=v, ncol=10, byrow = TRUE)
head(mx)

# Now that we have the matrix, "apply" applies a function to
# each row of the matrix, and returns the result as a vector
BYROW <- 1
row.means <- apply(X = mx, MARGIN = BYROW, FUN = mean)
head(row.means)

# Let's see the distribution of the row means.  It approximates
# a normal distribution, as predicted by the Central Limit Theorem.
hist(row.means, col="brown", density=20, freq = FALSE)
temp.x <- seq(0,6,.01)
lines(temp.x, dnorm(temp.x, mean=mean(row.means), sd=sd(row.means)))
```

Use "apply" to apply a function to every column of a matrix
```{r}
# Let's create a simple matrix and get the sum of each column
simple.mx <- matrix(data=c(1,2,3,
                           1000,1200,1100,
                           34,18,22),ncol = 3,byrow = FALSE)
colnames(simple.mx) <- c("subject","income","age")
BYCOL <- 2
(col.sums <- apply(X=simple.mx, MARGIN = BYCOL, FUN = sum))
# Let's show the matrix with its sums at the end of each column
rbind(simple.mx, col.sums)
```

Use "lapply" on each element of a list
```{r}
# First, we need a list
lst <- list()
lst$element1 <- (data.frame(x=1:3, y=4:6, z=c(NA,NA,NA)))
lst$element2 <- 1:100
(lr <- lapply(lst, sum, na.rm=TRUE))

```

Do the same thing with "sapply"; it will attempt to
create a vector of the results
```{r}
(lsr <- sapply(lst, sum, na.rm=TRUE))
```

"sapply" will also attempt to create a matrix if function
returns more than one value
```{r}
my.func <- function(o) {
  # This function returns column sum and the count of elements
  c(sum(o), NROW(o))
}
res <- sapply(lst, my.func)
rownames(res) <- c("sums","elements")
res

```

"tapply" applies a function to groups of a vector, or more
usefully, to grouped elements of a dataframe
```{r}
# Create a dataframe with groups
(df.pops <- read.csv(file = "vt_nh_counties.csv"))
# Use the population as the numeric field to group,
# and the state as the item to group on.
(state.pops <- tapply(X=df.pops$Population, INDEX = df.pops$State, FUN = sum))
```

The "by" function applies a function to groups of rows in a dataframe
```{r}
# use the same data frame
# XXX THIS IS NOT THE BEST EXAMPLE
by(data = df.pops, INDICES = df.pops$State, FUN = function(df) {lm})
```